Quantitative Evaluation of Touch-Screen Performance with Transition Matrix Measures

### 2015-09-14 flux-entropy-paper/flux-entropy-paper.tex
two faceted plots for beginning middle end

Suppose that each participant's activity in a collaborative
interaction can at any given time be classified as belonging to one of
$m$ distinct \emph{states}. Over the course of the activity, each
participant $X$ generates a length $N$ sequence of these interaction
states
\begin{equation}
 X_n \hskip 2em n = 1, \ldots, N
\end{equation}
The transitions between these interaction states can be summarised in
a transition matrix (TM) as if the sequence $X_n$ were a first-order Markov
chain. Let $N_{ij}$ be the number of times that state $j$ follows
state $i$. Each participant's TM $P$, will then be an
$m \times m$ matrix with entries
\begin{equation}
  p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
\end{equation}
Each entry of $P$ will be such that the transition probability from
state $i$ to state $j$ is given by the entry in the $i^{th}$ row and
the $j^{th}$ column. The transition activity of the whole ensemble can
be calculated by averaging the TM for each performer.
The matrix can be normalised so that the sum of all elements is equal
to one.
% (rather than the sum of each row being one, as for a right-stochastic matrix)

@@

To compare the TMs of different interactions we use two high-level
quantities, ``flux'' and ``entropy''. The flux of a sequence is
defined as the ratio of change transitions (e.g. $A \rightarrow B$
where $A \neq B$), to self transitions (e.g. $A \rightarrow A $). The
flux of a TM $P$ is given by
\begin{equation}
  \mathrm{flux}(P) = 1 - \mathrm{tr}(P)
\end{equation}
where the trace $\mathrm{tr}(P)$ is the sum of the diagonal entries of
$P$.
% Where $\|P\|_1 = \sum_{i,j}|p_{ij}|$ is the element-wise 1-norm of $P$
% and $\mathrm{diag}(P)$ is the vector of the main diagonal entries of
% $P$.
Intuitively, flux is a measure of how frequently the participant/group
changes state. Flux returns a value in the interval $[0,1]$ and will
return 0 when participants never change state, and 1 when participants
never stay in the same state for two measurements in a row.

@@

\section{Making sense of two year's worth of touch-screen
  interactions}

@@

Over a period of more than two years (April 2013 - June 2015), we have
collected data from $150$ group musicmaking sessions where
participants used musical touch-screen interfaces on Apple iPad
devices. These interfaces all shared a common mode of interaction
where the majority of the screen could be used for free form touch
interactions; tapping produced short sounds, while swiping or swirling
produced long sounds with volume proportional to the velocity of the
moving touch point. Through all of these performances every touch
interaction---every touch down, dragged and up---was timestamped and
logged to disk for later analysis

% TODO say something about how it was both your research project and
% your creative practice. That's why it encompassed all these
% different contexts.

Each session took place in a number of contexts including rehearsals,
performances, lab study sessions, classes, and casual demonstrations,
some of which are shown in Figure
\ref{fig:three-performance-contexts}. Participants performed free
improvisations as well as written compositions and performed on iPads
alone or using setups of iPads and acoustic percussion instruments.
The median length of these sessions was 7m29s and the median number of
participants was four, other descriptive statistics about the sessions
is shown in Table \ref{corpus-table}. Each session was coded with
metadata regarding the session context (whether it was a performance,
rehearsal, study, demonstration, or data collection session), the
session type (either improvisation or composition), and the
instruments used (iPads alone, or iPads with acoustic percussion). The
distribution of recorded sessions by context and type is shown in
Figure \ref{fig:count-data}.

@@

To analyse the gestural interactions present in each of these
sessions, the sessions were automatically coded according to a
vocabulary of nine simple, continuous, touch-screen gestures. Examples
of these gestures was generated by an expert user in formal process
where each gesture was performed for 1-minute and feature vectors of
discriptive statistics on these examples was used to train a Random
Forest~\cite{Breiman:2001kx} classifier\footnote{The classifier used
  was from Python's Scikit-learn package~\cite{scikit-learn}.}. The
classifier was applied to 5-second windows of the touch-data each
second provide a chain of gesture-states for each participant in each
session. These chains can be used to create $9 /times 9$
gesture-transition matrices for individual performers, and the whole
ensemble in each session.

@@

The distribution of flux and entropy with respect to session contexts,
types and instrumentation are shown in Figure
\ref{fig:flux-entropy-boxplot}. A three-way ANOVA procedure was
performed on the data set to find whether the metatdata could
significantly predict the entropy and flux values.
% TODO we haven't explained what ``the 95 performative sessions'' are,
% have we?
In the following analysis, only the 95 performative sessions were
considered (i.e.~not the demonstration or data collection contexts).

@@

\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.4\linewidth}
        \centering
          \includegraphics[width=\linewidth]{figures/type-section-flux}
          \includegraphics[width=\linewidth]{figures/type-section-entropy}
          \caption{Performances divided by performance type. \label{fig:type-section-flux-entropy}}
    \end{subfigure}%
    ~ 
    \begin{subfigure}[t]{0.4\linewidth}
        \centering
          \includegraphics[width=\linewidth]{figures/context-section-flux}
          \includegraphics[width=\linewidth]{figures/context-section-entropy}
          \caption{Sessions divided by performance context. \label{fig:context-section-flux-entropy}}
    \end{subfigure}
    \caption{The flux and entropy of different interaction sessions
      divided into beginning, middle, and end.}
\end{figure*}

>>>>

Suppose that each participant's activity in a collaborative
interaction can at any given time be classified as belonging to one of
$m$ distinct \emph{states}. Over the course of the activity, each
participant $X$ generates a length $N$ sequence of these interaction
states
\begin{equation}
 X_n \hskip 2em n = 1, \ldots, N
\end{equation}
The transitions between these interaction states can be summarised in
a transition matrix (TM) as if the sequence $X_n$ were a first-order Markov
chain. Let $N_{ij}$ be the number of times that state $j$ follows
state $i$. Each participant's TM $P$, will then be an
$m \times m$ matrix with entries
\begin{equation}
  p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
\end{equation}
Each entry of $P$ will be such that the transition probability from
state $i$ to state $j$ is given by the entry in the $i^{th}$ row and
the $j^{th}$ column. The transition activity of the whole ensemble can
be calculated by averaging the TM for each performer.
The matrix can be normalised so that the sum of all elements is equal
to one.

@@

To compare the TMs of different interactions we use two high-level
quantities, ``flux'' and ``entropy''. The flux of a sequence is
defined as the ratio of change transitions (e.g. $A \rightarrow B$
where $A \neq B$), to self transitions (e.g. $A \rightarrow A $). The
flux of a TM $P$ is given by
\begin{equation}
  \mathrm{flux}(P) = 1 - \mathrm{tr}(P)
\end{equation}
where the trace $\mathrm{tr}(P)$ is the sum of the diagonal entries of
$P$. Intuitively, flux is a measure of how frequently the participant/group
changes state. Flux returns a value in the interval $[0,1]$ and will
return 0 when participants never change state, and 1 when participants
never stay in the same state for two measurements in a row.

@@

% \section{Making sense of two year's worth of touch-screen
%   interactions}
\section{A corpus of touch-screen interaction}

@@

Over a period of more than two years (April 2013 - June 2015), we have
collected data from $150$ group music-making sessions where
participants used musical touch-screen interfaces on Apple iPad
devices. These interfaces all shared a common mode of interaction
where the majority of the screen could be used for free form touch
interactions; tapping produced short sounds, while swiping or swirling
produced long sounds with volume proportional to the velocity of the
moving touch point. Through all of these performances every touch
interaction---every touch down, dragged and up---was timestamped and
logged to disk for later analysis.

These sessions were part of a long-term research project into music
touch-screen interaction, as well as an ongoing creative practice. As
a result, the sessions had a variety of contexts including rehearsals,
performances, lab study sessions, classes, and casual demonstrations,
some of which are shown in Figure
\ref{fig:three-performance-contexts}. Participants performed free
improvisations as well as written compositions and performed on iPads
alone or using setups of iPads and acoustic percussion instruments.
The median length of these sessions was 7m29s and the median number of
participants was four, other descriptive statistics about the sessions
is shown in Table \ref{corpus-table}. Each session was coded with
metadata regarding the session context (whether it was a performance,
rehearsal, study, demonstration, or data collection session), the
session type (either improvisation or composition), and the
instruments used (iPads alone, or iPads with acoustic percussion). The
distribution of recorded sessions by context and type is shown in
Figure \ref{fig:count-data}.

@@

To analyse the gestural interactions present in each of these
sessions, the sessions were automatically coded according to a
vocabulary of nine simple, continuous, touch-screen gestures. Examples
of these gestures was generated by an expert user in formal process
where each gesture was performed for 1-minute and feature vectors of
discriptive statistics on these examples was used to train a Random
Forest~\cite{Breiman:2001kx} classifier\footnote{The classifier used
  was from Python's Scikit-learn package~\cite{scikit-learn}.}. The
classifier was applied to 5-second windows of the touch-data each
second provide a chain of gesture-states for each participant in each
session. These chains can be used to create $9 \times 9$
gesture-transition matrices for individual performers, and the whole
ensemble in each session.

@@

The distribution of flux and entropy with respect to session contexts,
types and instrumentation are shown in Figure
\ref{fig:flux-entropy-boxplot}. A three-way ANOVA procedure was
performed on the data set to find whether the metatdata could
significantly predict the entropy and flux values.
% TODO we haven't explained what ``the 95 performative sessions'' are,
% have we?
In the following analysis, only the 95 performative sessions were
considered, excluding the demonstration and data-collection contexts.

@@

\begin{figure*}[t!]
    \centering
    \begin{subfigure}[t]{0.5\linewidth}
        \centering
          \includegraphics[width=\linewidth]{figures/type-section-flux-entropy}
          \caption{Performances divided by performance type. \label{fig:type-section-flux-entropy}}
    \end{subfigure}%
    ~ 
    \begin{subfigure}[t]{0.5\linewidth}
        \centering
          \includegraphics[width=\linewidth]{figures/context-section-flux-entropy}
          \caption{Sessions divided by performance context. \label{fig:context-section-flux-entropy}}
    \end{subfigure}
    \caption{The flux and entropy of different interaction sessions
      divided into beginning, middle, and end.}
\end{figure*}
